The Markov link method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-17 paper/paper.tex
Dave edits

\title{The Markov Link Method: a nonparametric approach to combine observations from multiple experiments}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski, \\Additional Authors To Be Determined}

@@

\begin{abstract}
We propose the Markov Link Method (\MLM), a way to synthesize experiments based on several measurement procedures.  For example, we might have different experiments measuring different aspects of human neurons, such as morphology, transcriptomics, or electrophysiology.  The \MLM{} uses this data to link these different aspects.  It answers the question ``We have observed the transcriptome of this cell -- what might its morphology have been?'' The method is nonparametric: it makes no assumptions about what the link looks like.  We evaluate the MLM on a pair of single-cell RNA techniques and gain new insight about the link between the two methods.
\end{abstract}

We here propose the Markov Link Method (\MLM), a technique to understand how different measurement procedures are related.  For our purposes, we assume each experiment has two aspects:
%
\begin{itemize}
    \item A \emph{selection procedure} defines how we select cells.  For example, one experiment might select only liver cells, and another might select only cells which express the Vip protein.  
    \item A \emph{measurement procedure} defines what we measure about each cell.  For example, one experiment might measure the transcriptomic profile of a cell, and another experiment might measure the cell morphology.
\end{itemize}
%
To synthesize experiments with different measurement procedures, we must understand how the measurement procedures are related.  For example, we would like to be able to answer questions such as ``We have measured the transcriptome of this cell -- what might its morphology have been?''  We formalize this idea through what we call the `measurement link,' $\pxy(y|x)$.  Specifically, let us say we obtain measurement result $x$ from one measurement procedure.  The number $\pxy(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.

We here propose the The Markov Link Method (\MLM), an algorithm that uses a collection of experiments to estimate the measurement link, \pxy.  In this paper we focus particularly on the case that we have experiments based on several selection procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The \MLM{} could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the \MLM.  It embodies a common situation in biological problems, such as is found in \cite{gouwens2018classification}.

@@

\caption{{\bf The Markov Link Method (MLM) links different measurement procedures by using a set of different selection procedures.}  In subfigure (a) we have neurons gathered using many different selection procedures.  We select half of the neurons and determine what type of neuron they are, using transcriptomic measurements.  We classify the second half of neurons based on a morphological properties.  The MLM gives a way to link these two types of measurement and reconcile the two notions of cell-type.  In subfigure (b) we present a plate diagram articulating the statistical assumption that makes it possible (cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram).  For each neuron in each group, this diagram considers the selection procedure ($\ell$), the transcriptomic type ($X$), and the morphological type ($Y$).  The assumption is that the transcriptomic type is sufficiently detailed to \emph{statistically isolate} the transcriptomic type from the selection procedure.  This is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to estimate the link, $\pxy$. \label{fig:bigpicture}}
}

@@

\section{The Markov Link Method}

@@

The Markov Link Method is a general way to estimate measurement link from a collection of experiments.  For concreteness, we will use experiments about human cells as a running example.  Let us say that we have gathered many human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the selection procedure used to obtain the specimen
    \item $X$, the cell's `transcriptomic type,' as measured through a transcriptomic analysis.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to link the transcriptomic and morphological perspectives on `cell-type,' as manifested through the $X$ and $Y$ measurement procedures.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of selection procedure.  In particular:

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$\mathbb{P}(Y|X,\ell)=\pxy(y|x)$ for each $\ell$
\end{center}
\vspace{.1in}
}}
\end{center}

Here $\mathbb{P}(Y|X,\ell)$ indicates the probability that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  The assumption states that that this probability actually does not depend upon $\ell$.  This assumption can be understood intuitively through a thought experiment.  Let us imagine we have attained perfect understanding of cellular biology.  Someone selects a cell using one of a set of selection procedures and measures the transcriptomic activity of the cell.  We are then told the cell's transcriptomic type (but we are not told which of the selection procedures was used to gather the specimen).  Using our perfect knowledge of the physical systems, we could then make predictions about the cell's morphological type.  Now we are told new information: we are told which selection procedure was used to gather the specimen.  Would this substantially change our predictions?  If so, the MLM assumption is violated.  However, if the transcriptomic measurement is sufficiently informative, learning the sampling strategy would not substantially change our predictions about the cell morphology.  As long as we can find a measurement procedure that is sufficiently informative in this way, we can apply the MLM.  A list of examples where the MLM might apply may be found in Section \ref{sec:examples}.

More formally, our task is to estimate the link $\pxy$ from a collection of experiments.  We suppose that for each selection procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $\DX$ and $\DY$.  The matrix $\DX$ tabulates the transcriptomic types found in specimens gathered with each selection procedure; $\DY$ does the same for morphological types.   For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses this data to produce two objects:
%
\begin{enumerate}
    \item $\hat \pxy(y|x)$ -- a point estimate for the link $\pxy(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link $\pxy(y|x)$ with high probability
\end{enumerate}
%
To estimate these, we must consider certain `nuisance' quantities.  The link is our primary target of study, but we cannot estimate the link without considering these other objects.  Our observable data, $\DX,\DY$, is governed by the following conditional distributions:
\begin{itemize}
    \item $\plx(x|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$
    \item $\ply(y|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have morphological type $y$
\end{itemize}
It is straightforward to estimate $\plx,\ply$ from $\DX,\DY$ using standard methods.  We would like to use these distributions to tell us something about our object of interest, the link $\pxy$.  Under the Markov Link Method assumption, this link is connected to $\plx,\ply$ through the equations
\begin{gather}\label{eq:mlmh}
\sum_x \plx(x|\ell)\pxy(y|x)=\ply(y|\ell) \qquad \forall y,\ell
\end{gather}
With these equations and knowledge of $\plx,\ply$, we can produce an estimate for the link: $\hat \pxy_{\plx,\ply}$.  If the equations are underdetermined, we can also produce bounds for the link.  We define these objects precisely in Appendix \ref{sec:mlmdetails}.  However, in practice we do not actually have exact knowledge of $\plx,\ply$.  We only have estimates from data.  We take a Bayesian perspective to account for our uncertainty in this estimation.  For our prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$.  Following the Bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge about $\plx,\ply$.  First, we have observed the data, $\DX,\DY$.  Second, we know from the MLM assumption that there exists \emph{some} value $\pxy$ such that Equation (\ref{eq:mlmh}) holds; let \MLMH{} denote the event that this holds.  By conditioning on our knowledge, we obtain the posterior distribution $\mathbb{P}(d\plx,d\ply | \DX,\DY,\MLMH)$.  Our point estimate $\hat \pxy$ is then defined as a Bayes estimator for $\hat \pxy_{\plx,\ply}$ with respect to this posterior:
\[
\hat \pxy(y|x) \triangleq \mathbb{E}\left[\hat \pxy_{\plx,\ply} | \DX,\DY,\MLMH \right]
               = \int \hat \pxy_{\plx,\ply} \mathbb{P}(d\plx,d\ply | \DX,\DY,\MLMH) 
\]
This estimate minimizes the expected squared risk error under the posterior distribution.  We can use similar strategies to produce the credible intervals $C_{x,y}$.  Exact details can be found in Appendix \ref{sec:mlmdetails}.  Code to compute $\hat \pxy(y|x)$ and $C_{x,y}$ is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

\section{Examples where the Markov Link Method may apply}

@@

The validity of the Markov Link Method assumption for a given situation should be closely contemplated.  Let us consider a few real-world examples where this assumption may apply.

@@

    \item Cancer treatment efficacy prediction.  Starting from in-vivo human cancers, many cell-lines have been cultured over the years.  These cell cultures live indefinitely on plates.  Many experiments have been performed to see how these cancer cells respond to treatment.  However, if a treatment works on a particular cultured cell-line, what can we say about whether a treatment will work on an actual in-vivo cancer inside a patient?  Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.  Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).  However, we know that cultured cell-lines look quite different from in-vivo cells (cf. \cite{imamura2015comparison,haibe2013inconsistency}).   These cell cultures are subject to quite different pressures, due to the fact that they survive on a plate instead of inside a human being.  The Markov Link method can leverage the common side-information together with separate transcriptome information to understand the correspondence between in-vivo and cultured cells.  If a particular drug is effective on a particular cultured cell-line, we can then look at the corresponding in-vivo transcriptomic profile.  If we find human cancers that match this profile, they might be good candidates for further research using this particular drug.  Here $\ell$ might indicate cancer location, $X$ might indicate transcriptomic expression of cultured cells, and $Y$ might indicate transcriptomic expression of in-vivo cells.  As the transcriptomic expression is much more informative than the cancer location, it is plausible that $X$ might be sufficient to explain any correlations between $\ell$ and $Y$.  Thus the MLM assumption may hold.

    \item Text/image correspondence.  Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).  There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.  Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.  The Markov Link Method suggests one way to use the more plentiful text-only and image-only data.  We can first apply classic machine learning techniques to get coarse labels for both kinds of data.  Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.  Here $\ell$ would indicate coarse labels such as ``cat'' or ``street scene.''  These labels could be derived from either images or text and can be trained in a supervised fashion.  $X$ would indicate the image and $Y$ would indicate a caption.  If the caption is largely determined by the picture $X$, the MLM assumption may hold.

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, selection procedures, and the link itself.  In each trial our `ground truth' link picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct selection procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link $\pxy(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link probabilities lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct selection procedures.   In subfigure (c) we focus on the case of exactly $4$ selection procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of selection procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat \pxy(y|x)$ should converge to the true link $\pxy(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $\pxy(y|x)$.
    \item[Conservative coverage] $\pxy(y|x) \in C_{x,y}$ with high probability.
\end{description}
In many cases, the MLM achieves all three of these conditions.  In other cases, we can show that an issue known as `identifiability' blocks even the possibility of estimator convergence or interval concentration.  However, even when interval concentration is impossible, the \MLM{} appears to do as well as it possibly can: the intervals $C_{x,y}$ close around the bounds of what can possibly be learned from the data we observe.  We have begun to develop a theory of the relevant issues, which we detail in Appendix \ref{sec:mlmdetails}.  We give a few of the interesting corner-cases and prove a few surprising results.  A complete story remains elusive.  The following two simulations give a good overview of the relevant issues:

@@

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct selection procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or selection procedures; they include the ground truth with high probability.   With many samples and selection procedures, the intervals are small and concentrated around the truth.  With fewer samples or selection procedures, the intervals are sometimes forced to be larger.  However, our uncertainty is not the same for every aspect of the link.  In some cases we are able to obtain a very tight credible interval for one particular value even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link itself?  In each trial of this simulation we use four selection procedures and our measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of selection procedures.  However, in our previous simulations we picked the link uniformly from its parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $\pxy(y|x)=\pxy(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen selection procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

\figgy{
\hfill{}\figgygr{.95}{../images/allenNlxy}\hfill{}
\caption{{\bf The input to the Markov Link Method: one experiment for each measurement procedure and each selection procedure.}  Here we show a portion of the real-world data to which we applied the MLM.  The Allen Institute gathered neurons from the visual cortex of mice, using a variety of cre/lox-based selection procedure (cf.\ \cite{tasic2017shared}).  Each strategy was designed to bring out different subpopulations of cells.  Neurons were measured to determine their `type,' using one of two procedures: Standard or Patch.  Standard recognizes 104 different types of neurons.  Patch has a coarser notion of cell-type, distinguishing only 10 types.  For each experiment, we tabulated the number of cells assigned to each type.  Above we show a subset of these results through heatmaps; the color of each square indicates the number of specimens found to have a particular type.  Using this kind of data, our task is to calibrate the two classification protocols.  That is, we want to be able to ask question of the following form: ``if a neuron is classified as being of type `Peri Kcnj8' by Standard, how might it have been classified by Patch?'' \label{fig:alleninput}}
}

@@

In future work, it would be interesting to use additional assumptions to help us estimate calibrations.  For example, as it stands the Markov Link Method will only yield narrow credible intervals if each measurement tool returns one of a finite number of measurement outcomes.  In some cases, we may believe that similar measurement values should have similar probabilities.  Such smoothness assumptions would make it possible to apply the MLM to measurement tools which can return a continuum of values.

@@

\section{Exact details of the Markov Link Method}

@@

There are three numerical problems which the MLM must solve.  Here we detail our method for solving each of them.

>>>>

\title{The Markov link method: a nonparametric approach to combine observations from multiple experiments}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski, \\Additional Authors To Be Determined}

@@

\begin{abstract}
We propose the Markov link method (\MLM), a way to synthesize experiments based on several measurement procedures.  For example, we might have different experiments measuring different aspects of human neurons, such as morphology, transcriptomics, or electrophysiology.  The \MLM{} uses this data find the  link between different aspects.  This link answers questions such as: ``We have observed the transcriptome of this cell -- what might its morphology have been?'' The method inspects how different \emph{measurement} procedures relate by using a collection of \emph{sample selection} procedures.   In doing so, it makes no assumptions about what the link looks like.  Instead, the \MLM{} relies on a conditional independence assumption.  We prove conditions under which the \MLM{} estimators are consistent, and simulations show that it gives accurate measures of uncertainty.  We evaluate the MLM on a pair of single-cell RNA techniques and gain new insight about the link between the two methods; we see aspects in which two different notions of `cell-type' are aligned, aspects in which they are not, and aspects about which we are unsure.  The \MLM{} suggests specific additional experiments which would help refine our understanding of the relationship between the two measurement procedures.
\end{abstract}

It is sometimes difficult to understand how to synthesize experiments which use different measurement procedures.  We assume each experiment has two aspects:
%
\begin{itemize}
    \item A \emph{selection procedure} defines how we select cells.  For example, one experiment might select only liver cells, and another might select only cells which express the Vip protein.  
    \item A \emph{measurement procedure} defines what we measure about each cell.  For example, one experiment might measure the transcriptomic profile of a cell, and another experiment might measure the cell morphology.
\end{itemize}
%
To synthesize experiments with different measurement procedures, we must understand how the measurement procedures are related.  For example, we would like to be able to answer questions such as ``We have measured the transcriptome of this cell -- what might its morphology have been?''  We formalize this idea through a conditional probability distribution we call the `measurement link,' $\pxy(y|x)$.  Specifically, let us say we obtain measurement result $x$ from one measurement procedure.  The number $\pxy(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.

We here propose the The Markov link method (\MLM), an algorithm that uses a collection of experiments to estimate the measurement link, \pxy.  In this paper we focus particularly on the case that we have experiments based on several selection procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The \MLM{} could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the \MLM.  It embodies a common situation in biological problems, such as is found in \cite{gouwens2018classification}.

@@

\caption{{\bf The Markov link method (MLM) links different measurement procedures by using a set of different selection procedures.}  In subfigure (a) we have neurons gathered using many different selection procedures.  We select half of the neurons and determine what type of neuron they are, using transcriptomic measurements.  We classify the second half of neurons based on a morphological properties.  The MLM gives a way to link these two types of measurement and reconcile the two notions of cell-type.  In subfigure (b) we present a plate diagram articulating the statistical assumption that makes it possible (cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram).  For each neuron in each group, this diagram considers the selection procedure ($\ell$), the transcriptomic type ($X$), and the morphological type ($Y$).  The assumption is that the transcriptomic type is sufficiently detailed to \emph{statistically isolate} the transcriptomic type from the selection procedure.  This is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to estimate the link, $\pxy$. \label{fig:bigpicture}}
}

@@

\section{The Markov link method}

@@

The Markov link method is a general way to estimate measurement links using a collection of experiments.  For concreteness, we will use experiments about human cells as a running example.  Let us say that we have gathered many human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the selection procedure used to obtain the cell.  For example, $\ell=1$ might correspond to a procedure that selects only cells exhibiting one proteomic marker, whereas $\ell=2$ might select cells with a different marker.
    \item $X$, the cell's `transcriptomic type,' as measured through a transcriptomic analysis.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
The \MLM{} gives us a way to link the transcriptomic and morphological perspectives on `cell-type,' as manifested through the $X$ and $Y$ measurement procedures.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''

The \emph{Markov link method assumption} states that these kinds of associations are unaffected by the choice of selection procedure.  That is, $\mathbb{P}(Y|X,\ell)=\pxy(y|x)$ for each $\ell$, where $\mathbb{P}(Y|X,\ell)$ indicates the probability that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  The assumption states that that this probability actually does not depend upon $\ell$.  The graphical model in figure \ref{fig:bigpicture} embodies precisely this assumption.  Intuitively, we can understand the assumption using a thought experiment.  Let us imagine we have attained perfect understanding of cellular biology.  Someone selects a cell using one of a set of selection procedures and measures the transcriptomic activity of the cell.  We are then told the cell's transcriptomic type (but we are not told which of the selection procedures was used to gather the specimen).  Using our perfect knowledge of the physical systems, we could then make predictions about the cell's morphological type.  Now we are told new information: we are told which selection procedure was used to gather the specimen.  Would this substantially change our predictions?  If so, the MLM assumption is violated.  However, if the transcriptomic measurement is sufficiently informative, learning the sampling strategy would not substantially change our predictions about the cell morphology.  As long as we can find a measurement procedure that is sufficiently informative in this way, we can apply the MLM.  A list of examples where the MLM might apply may be found in Section \ref{sec:examples}.

The purpose of the \MLM{} is to use data and the \MLM{} assumption to estimate the link $\pxy$ from a collection of experiments.  We suppose that for each selection procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $\DX$ and $\DY$.  The matrix $\DX$ tabulates the transcriptomic types found in specimens gathered with each selection procedure; $\DY$ does the same for morphological types.   For each transcriptomic type $x$ and morphological type $y$, the Markov link method uses this data to produce two objects:
%
\begin{enumerate}
    \item $\hat \pxy(y|x)$ -- a point estimate for the link $\pxy(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link $\pxy(y|x)$ with high probability
\end{enumerate}
%
To estimate these, we must consider certain `nuisance' quantities.  The link is the primary target of study, but we cannot estimate the link without considering these other objects.  The observable data, $\DX,\DY$, is governed by the following conditional distributions:
\begin{itemize}
    \item $\plx(x|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$
    \item $\ply(y|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have morphological type $y$
\end{itemize}
It is straightforward to estimate $\plx,\ply$ from $\DX,\DY$ using standard methods.  We would like to use these distributions to tell us something about the the link, $\pxy$.  Under the Markov link method assumption, this link is connected to $\plx,\ply$ through the equations
\begin{gather}\label{eq:mlmh}
\sum_x \plx(x|\ell)\pxy(y|x)=\ply(y|\ell) \qquad \forall y,\ell
\end{gather}
With these equations and knowledge of $\plx,\ply$, we can produce an estimate for the link: $\hat \pxy_{\plx,\ply}$.  If the equations are underdetermined, we can also produce bounds for the link.  We define these objects precisely in Appendix \ref{sec:mlmdetails}.  However, in practice we do not actually have exact knowledge of $\plx,\ply$.  We only have estimates from data.  We take a Bayesian perspective to account for uncertainty in this estimation.  For prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$.  Following the Bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge about $\plx,\ply$.  First, we have observed the data, $\DX,\DY$.  Second, we know from the MLM assumption that there exists \emph{some} value $\pxy$ such that Equation (\ref{eq:mlmh}) holds; let \MLMH{} denote the event that this holds.  By conditioning on this knowledge, we obtain the posterior distribution $\mathbb{P}(\plx,\ply | \DX,\DY,\MLMH)$.  The \MLM{} point estimate $\hat \pxy$ is then defined as a Bayes estimator for $\hat \pxy_{\plx,\ply}$ with respect to this posterior:
\[
\hat \pxy(y|x) \triangleq \mathbb{E}\left[\hat \pxy_{\plx,\ply} | \DX,\DY,\MLMH \right]
               = \int \hat \pxy_{\plx,\ply} \mathbb{P}(\plx,\ply | \DX,\DY,\MLMH) d\plx,d\ply
\]
This estimate minimizes the expected squared risk error under the posterior distribution.  We can use similar strategies to produce the credible intervals $C_{x,y}$.  Exact details can be found in Appendix \ref{sec:mlmdetails}.\footnote{Code to compute $\hat \pxy(y|x)$ and $C_{x,y}$ is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.}

@@

\subsection{Examples}

@@

The \MLM{} is a general approach for synthesizing experiments, which requires that the Markov link method assumption holds.  Here are some examples where the \MLM{} could be applied:

@@

    \item Cancer treatment efficacy prediction.  Starting from in-vivo human cancers, many cell-lines have been cultured over the years.  These cell cultures live indefinitely on plates.  Many experiments have been performed to see how these cancer cells respond to treatment.  However, if a treatment works on a particular cultured cell-line, what can we say about whether a treatment will work on an actual in-vivo cancer inside a patient?  Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.  Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).  However, we know that cultured cell-lines look quite different from in-vivo cells (cf. \cite{imamura2015comparison,haibe2013inconsistency}).   These cell cultures are subject to quite different pressures, due to the fact that they survive on a plate instead of inside a human being.  The Markov link method can leverage the common side-information together with separate transcriptome information to understand the correspondence between in-vivo and cultured cells.  If a particular drug is effective on a particular cultured cell-line, we can then look at the corresponding in-vivo transcriptomic profile.  If we find human cancers that match this profile, they might be good candidates for further research using this particular drug.  Here $\ell$ might indicate cancer location, $X$ might indicate transcriptomic expression of cultured cells, and $Y$ might indicate transcriptomic expression of in-vivo cells.  As the transcriptomic expression is much more informative than the cancer location, it is plausible that $X$ might be sufficient to explain any correlations between $\ell$ and $Y$.  Thus the MLM assumption may hold.

    \item Text/image correspondence.  Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).  There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.  Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.  The Markov link method suggests one way to use the more plentiful text-only and image-only data.  We can first apply classic machine learning techniques to get coarse labels for both kinds of data.  Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.  Here $\ell$ would indicate coarse labels such as ``cat'' or ``street scene.''  These labels could be derived from either images or text and can be trained in a supervised fashion.  $X$ would indicate the image and $Y$ would indicate a caption.  If the caption is largely determined by the picture $X$, the MLM assumption may hold.

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, selection procedures, and the link itself.  In each trial the `ground truth' link is picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct selection procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link $\pxy(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link probabilities lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct selection procedures.   In subfigure (c) we focus on the case of exactly $4$ selection procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of selection procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov link method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat \pxy(y|x)$ should converge to the true link $\pxy(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $\pxy(y|x)$.
    \item[Conservative coverage] $\pxy(y|x) \in C_{x,y}$ with high probability.
\end{description}
In many cases, the MLM achieves all three of these conditions.  In other cases, we can show that an issue known as `identifiability' blocks even the possibility of estimator convergence or interval concentration.  However, even when interval concentration is impossible, the \MLM{} appears to do as well as it possibly can: the intervals $C_{x,y}$ close around the bounds of what can possibly be learned from the data we observe.  We have begun to develop a theory of the relevant issues, which we detail in Appendix \ref{sec:mlmdetails}.  We give a few of the interesting corner-cases and prove a few surprising results.  A complete story remains elusive.  The following two simulations give a good overview of the relevant issues:

@@

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct selection procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or selection procedures; they include the ground truth with high probability.   With many samples and selection procedures, the intervals are small and concentrated around the truth.  With fewer samples or selection procedures, the intervals are sometimes forced to be larger.  However, the uncertainty is not the same for every aspect of the link.  In some cases we are able to obtain a very tight credible interval for one particular value even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link itself?  In each trial of this simulation we use four selection procedures and the measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of selection procedures.  However, in previous simulations we picked the link uniformly from its parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $\pxy(y|x)=\pxy(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen selection procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

\figgy{
\hfill{}\figgygr{.95}{../images/allenNlxy}\hfill{}
\caption{{\bf The input to the Markov link method: one experiment for each measurement procedure and each selection procedure.}  Here we show a portion of the real-world data to which we applied the MLM.  The Allen Institute gathered neurons from the visual cortex of mice, using a variety of cre/lox-based selection procedure (cf.\ \cite{tasic2017shared}).  Each strategy was designed to bring out different subpopulations of cells.  Neurons were measured to determine their `type,' using one of two procedures: Standard or Patch.  Standard recognizes 104 different types of neurons.  Patch has a coarser notion of cell-type, distinguishing only 10 types.  For each experiment, we tabulated the number of cells assigned to each type.  Above we show a subset of these results through heatmaps; the color of each square indicates the number of specimens found to have a particular type.  Using this kind of data, the task is to calibrate the two classification protocols.  That is, we want to be able to ask question of the following form: ``if a neuron is classified as being of type `Peri Kcnj8' by Standard, how might it have been classified by Patch?'' \label{fig:alleninput}}
}

@@

In future work, it would be interesting to use additional assumptions to help us estimate calibrations.  For example, as it stands the Markov link method will only yield narrow credible intervals if each measurement tool returns one of a finite number of measurement outcomes.  In some cases, we may believe that similar measurement values should have similar probabilities.  Such smoothness assumptions would make it possible to apply the MLM to measurement tools which can return a continuum of values.

@@

\section{Exact details of the Markov link method}

@@

There are three numerical problems which the MLM must solve.  Here we detail the \MLM{} method for solving each of them.
